---
title: "Mind Matters: A Data-driven Examination of College Students' Mental Health"
subtitle: "Statistical Inferecence and Learining project"
author: "Giovanni Costa - 880892"
date: "AY 2023/24"
geometry: "left=1cm,right=2cm,top=1cm,bottom=1cm"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      smooth_scroll: false
    fig_caption: yes
    theme: flatly
    highlight: pygments
    css: "assets/css/styles.css"
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction
This project aims to delve into the pressing issue of mental health among college students, a topic of significant importance in today's fast-paced, high-stress academic environment. By utilizing a data-driven approach, the analysis seeks to provide a comprehensive understanding of the mental health landscape in universities, with a particular focus on psychiatric symptoms, suicidal ideation, and other associated health problems.

The dataset, that can be retrieved <a href="https://psycharchives.org/en/item/5d25cf8a-7910-4579-9e39-688c6bbec43a">here</a> contains 27014 objects with 16 features. The first six are categorical information about the general individual condition, instead the others are numeric variables related to the mental health of the students, where the higher score indicates more serious symptoms.<br/>
A detailed description of these features is provided below:

1. **gender**: 0=Female, 1=Male

2. **whether_only_child**: 0=No, 1=Yes

3. **birth_place**: 0=Countryside, 1=Town, 2=SmallCity, 3=MediumToLargeCities

4. **family_economic_status**: 0=ExtremelyPoor, 1=Poor, 2=Average, 3=Good, 4=Rich

5. **major**: 0=Liberal, 1=Science, 2=Art

6. **grade**: 0=Postgraduate, 1=UndergraduateGradeFive, 2=Junior, 3=Sophomore, 4=Freshman, 5=Senior

7. **psychiatric_symptoms**: range from 4 to 16

8. **suicide**: range from 4 to 16

9. **dependence**: range from 4 to 16

10. **impulsivity**: range from 4 to 16                                                         

11. **compulsion**: range from 4 to 16

12. **sleeping_disturbance**: range from 4 to 16

13. **internet_addiction**: range from 5 to 20

14. **hostile_aggression**: range from 4 to 16

15. **self_injury_behaviors:** range from 4 to 16

16. **eating_problems**: range from 4 to 16

The purposes of this analysis are inspecting the possible hidden patterns present in the data and predicting the risk of observing a suicide, modeled for convenience as a binary variable. 

# Prerequirements
```{r results='hide', message=FALSE, warning=FALSE}
set.seed(123) # set pseudorandom generator for reprocucibility

# Load all the packages and install them if they are not present
requirements <- c("summarytools", "MASS", "effects", "pROC", "mgcv", 
                  "glmnetUtils", "e1071", "class",
                  "stringr", "ggplot2", "reshape2", "scales",  "comprehenr", "randomForest", "dplyr")

for (library_name in requirements){
  if (!require(library_name, character.only = TRUE)) {
    install.packages(library_name, repos = "https://cloud.r-project.org")
    library(library_name, character.only = TRUE)
  }
}
# Import user defined functions
source("utils.R")
# library(styler)
# style_file("analysis_report.Rmd")
```


# First overview and data engineering

In this section, exploratory data analysis (EDA) is performed to understand the structure of the dataset and identify any potential issues that need to be addressed. Also, data engineering is applied to transform the variables into the appropriate format.
```{r}
# Read the dataset
df <- read.csv("data/mental_health_data.csv")
# Remove the column representing the row index
df <- df[, -1]

# Rename existing columns
colnames(df)[colnames(df) == "whether.only.child"] <- "whether_only_child"
colnames(df)[colnames(df) == "birth.place"] <- "birth_place"
colnames(df)[colnames(df) == "family.economic.status"] <- "family_economic_status"
colnames(df)[colnames(df) == "psychiatric.symptoms"] <- "psychiatric_symptoms"
colnames(df)[colnames(df) == "sleeping.disturbance"] <- "sleeping_disturbance"
colnames(df)[colnames(df) == "internet.addiction"] <- "internet_addiction"
colnames(df)[colnames(df) == "hostile.aggression"] <- "hostile_aggression"
colnames(df)[colnames(df) == "self.injury.behaviors"] <- "self_injury_behaviors"
colnames(df)[colnames(df) == "eating.problems"] <- "eating_problems"


# Transforming variables into factors
df$gender <- factor(df$gender,
  levels = c(0, 1),
  labels = c("Female", "Male")
)
df$whether_only_child <- factor(df$whether_only_child,
  levels = c(0, 1),
  labels = c("No", "Yes")
)
df$birth_place <- factor(
  df$birth_place,
  levels = 0:3,
  labels = c("Countryside", "Town", "SmallCity", "MediumToLargeCities")
)
df$family_economic_status <- factor(
  df$family_economic_status,
  levels = 0:4,
  labels = c("ExtremelyPoor", "Poor", "Average", "Good", "Rich")
)
df$major <- factor(df$major,
  levels = 0:2,
  labels = c("Liberal", "Science", "Art")
)
df$grade <- factor(
  df$grade,
  levels = 0:5,
  labels = c(
    "Postgraduate",
    "UndergraduateGradeFive",
    "Junior",
    "Sophomore",
    "Freshman",
    "Senior"
  )
)

general_cols <- c("gender", "whether_only_child", "birth_place", "family_economic_status", "major", "grade")
symptoms_cols <- c("psychiatric_symptoms", "dependence", "impulsivity", "compulsion", "sleeping_disturbance", "internet_addiction", "hostile_aggression", "self_injury_behaviors", "eating_problems")
```

```{r}
print_summary_custom(df)
```

```{r}
# "Dataset dimensions:"
dim(df)
# "N. of missing values:"
sum(is.na(df))
# "N. of duplicated rows:"
sum(duplicated(df))
# "Example of some objects:"
head(df, 3)
```

From the above summary it can be seen that all the column have the right format now. Also can be observed that there are no missing values but some duplicate rows are present. Duplicates in this case probably refers to students that have the same profile conditions and that have replayed in the same way to the survey for building the dataset. It's choosen to keep these rows to mantain the original dataset semantics.

5 is the value of the 5 3rd quantile, 8 is more extreme and seems reasonable because we want to take into account just more serious symptoms

```{r}
numeric_suicide <- df$suicide
scaled_suicide <- numeric_suicide - min(numeric_suicide)
factor_suicide <- as.factor(df$suicide>8)

df$suicide <- factor_suicide
```


# Exploratory Data Analysis

## Frequency
```{r fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
for (col in general_cols){
  plot_freq_by_category(df, col, "suicide")
}
```

```{r fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
for (col in symptoms_cols){
  plot_freq_by_category(df, col, "suicide")
}
```


## Numeric variables
```{r}
cor_matrix <- cor(df[, symptoms_cols])   # Calculate correlations
melted_cor_matrix <- melt(cor_matrix)  # Reshape for ggplot2

ggplot(data = melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1,1), name = "Pearson\nCorrelation") +
  labs(title = "Heatmap of Correlation Matrix") +   # Add title
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),              # Remove x-axis title
    axis.title.y = element_blank(),              # Remove y-axis title
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),  # Adjust x-axis text
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  coord_fixed()
```

## Interaction terms
```{r}
df$suicide <- numeric_suicide
```

```{r message=FALSE, warning=FALSE}
ggplot(df, aes(x = psychiatric_symptoms, y = suicide, color = gender)) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE, aes(group = gender)) +
  labs(title = "Suicide vs psychiatric symptoms by gender interaction",
       x = "Psychiatric symptoms",
       y = "Suicide")
```

```{r message=FALSE, warning=FALSE}
ggplot(df, aes(x = eating_problems, y = suicide, color = family_economic_status)) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE, aes(group = family_economic_status)) +
  labs(title = "Suicide vs eating problems by family economic status interaction",
       x = "Eating problems",
       y = "Suicide")

```

```{r message=FALSE, warning=FALSE}
ggplot(df, aes(x = sleeping_disturbance, y = suicide, color = major)) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE, aes(group = major)) +
  labs(title = "Suicide vs sleeping disturbance by major interaction",
       x = "Sleeping disturbance",
       y = "Suicide")
```

```{r}
df$suicide <- factor_suicide
```


# Models developement
```{r}
idx <- sample(nrow(df), 0.80*nrow(df))
df_train <- df[idx,]
df_test <- df[-idx,]

# Training split dimension
dim(df_train)
# Testing split dimension
dim(df_test)
```

```{r}
mod_full <- glm(suicide ~ ., data=df_train, family="binomial")
summary(mod_full)
```

## Manual selection
```{r}
mod_manual <- glm(suicide ~ family_economic_status + psychiatric_symptoms +impulsivity +self_injury_behaviors, data=df_train, family="binomial")
summary(mod_manual)

```

## Stepwise regression
```{r}
mod_zero <- glm(suicide ~ 1, data=df_train, family="binomial")
mod_step <- stepAIC(mod_zero, direction = "both", scope = list(upper = formula(mod_full)), trace=0)
summary(mod_step)
```


## Lasso
```{r}
mod_lasso <- cv.glmnet(suicide ~ ., data = df_train, family = "binomial", alpha = 1)
coef_lasso <- coef(mod_lasso, s=mod_lasso$lambda.1se)[,1]
coeff_lasso_idx <- which(coef_lasso != 0)
print(cbind(coeff_lasso_idx, coef_lasso[coeff_lasso_idx])[,2])

plot(mod_lasso$glmnet.fit, xvar="lambda")
abline(v = log(mod_lasso$lambda.1se), lty = "dashed")

plot(mod_lasso)
abline(v = log(mod_lasso$lambda.1se), lty = "dashed")
```

## LDA
```{r}
mod_lda <- lda(suicide~., data=df_train)
```


## QDA
```{r}
mod_qda <- qda(suicide~., data=df_train)
```

## Naive Bayes
```{r}
mod_nb <- naiveBayes(suicide~., data = df_train)
```


## KNN
```{r}
x.train <- model.matrix(~ ., data=subset(df_train, select = -suicide))[, -1]
x.test <- model.matrix(~ ., data=subset(df_test, select = -suicide))[, -1]

k_max <- 3
rates <- double(k_max)
for (i in 1:k_max) {
  tmp <- knn(train = x.train, test = x.test, cl = df_train$suicide, k = i)
  rates[i] <- mean(tmp == df_test$suicide)
}
plot(x = (1:k_max), y = rates, xlab = "k", ylab = "Accuracy", type = "l")

best_k <- which.max(rates)
```

## Random Forest
```{r}
mod_rf <- randomForest(suicide ~ ., data=df_train, importance=TRUE, ntree=5)
print(mod_rf)

plot(mod_rf)
varImpPlot(mod_rf)
```


## Different model definition
```{r}
# FOR LATER
df$scaled_suicide <- scaled_suicide
df <- subset(df, select=-suicide)
df_train <- df[idx,]
df_test <- df[-idx,]
mod_test <- glm(cbind(scaled_suicide, (max(scaled_suicide) - scaled_suicide)) ~
             .,
           family = "binomial",
           data = df_train)
table(
 preds = predict(mod_test, newdata = df_test, type = "response"),
 true = df_test$scaled_suicide
)
```

# Model comparison
```{r}
probs_stepwise <- predict(mod_step, newdata = df_test, type = "response")
roc_stepwise <- roc(df_test$suicide ~ probs_stepwise, plot = TRUE, print.auc = TRUE)
coords_stepwise <- coords(roc_stepwise, x = "best", ret = c("threshold", "specificity", "sensitivity", "accuracy","tn","tp","fn","fp"))
print(coords_stepwise)
```

```{r}
probs_lasso <- predict(mod_lasso, newdata = df_test, s = mod_lasso$lambda.1se, type = "response")
roc_lasso <- roc(df_test$suicide ~ as.vector(probs_lasso), plot = TRUE, print.auc = TRUE)
coords_lasso <- coords(roc_lasso, x = "best", ret = c("threshold", "specificity", "sensitivity", "accuracy","tn","tp","fn","fp"))
print(coords_lasso)
```

```{r}
probs_lda <- predict(mod_lda, newdata = df_test, )
roc_lda <- roc(df_test$suicide ~ probs_lda$posterior[, 2], plot = TRUE, print.auc = TRUE)
coords_lda <- coords(roc_lda, x = "best", ret = c("threshold", "specificity", "sensitivity", "accuracy","tn","tp","fn","fp"))
print(coords_lda)
```

```{r}
probs_qda <- predict(mod_qda, newdata = df_test, )
roc_qda <- roc(df_test$suicide ~ probs_qda$posterior[, 2], plot = TRUE, print.auc = TRUE)
coords_qda <- coords(roc_qda, x = "best", ret = c("threshold", "specificity", "sensitivity", "accuracy","tn","tp","fn","fp"))
print(coords_qda)
```

```{r}
probs_nb <- predict(mod_nb, newdata = df_test, type = "raw")
roc_nb <- roc(df_test$suicide ~ probs_nb[, 2], plot = TRUE, print.auc = TRUE)
coords_nb <- coords(roc_nb, x = "best", ret = c("threshold", "specificity", "sensitivity", "accuracy","tn","tp","fn","fp"))
print(coords_nb)
```

```{r}
probs_knn <- knn(train = x.train, test = x.test, cl = df_train$suicide, k = best_k)
conf_matrix_knn <- table(df_test$suicide, probs_knn)
# Calculate True Positives (TP), True Negatives (TN), False Positives (FP), False Negatives (FN)
TP_knn <- conf_matrix_knn[2, 2]  # True Positives
TN_knn <- conf_matrix_knn[1, 1]  # True Negatives
FP_knn <- conf_matrix_knn[1, 2]  # False Positives
FN_knn <- conf_matrix_knn[2, 1]  # False Negatives

# Specificity (True Negative Rate)
specificity <- TN_knn / (TN_knn + FP_knn)
# Sensitivity (Recall or True Positive Rate)
sensitivity <- TP_knn / (TP_knn + FN_knn)
# Accuracy
accuracy <- (TP_knn + TN_knn) / sum(conf_matrix_knn)
coords_knn <- data.frame(threshold = NA, specificity = specificity, sensitivity = sensitivity, accuracy = accuracy, tn = TN_knn, tp = TP_knn, fn = FN_knn, fp = FP_knn)
```


```{r}
# allEffects((glm(suicide ~ gender +family_economic_status, data=df, family="binomial")))

# residualPlots(mod_zero)
# qqPlot(residuals(mod_full))
# influenceIndexPlot(mod_full, vars=c("Cook"))
# vif(model)


# plot(allEffects(mod3), ylab = "Probability of Survival", rescale.axis = FALSE)
# mod_zero <- glm(suicide ~ 1, data=df, family="binomial")
```

