---
title: "Mind Matters: A Data-driven Examination of College Students' Mental Health"
subtitle: "Statistical Inferecence and Learining project"
author: "Giovanni Costa - 880892"
date: "AY 2023/24"
geometry: "left=3cm,right=3cm,top=1cm,bottom=1cm"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      smooth_scroll: false
    fig_caption: yes
    theme: flatly
    highlight: pygments
    css: "assets/css/styles.css"
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

1. **gender**: 0=female, 1=male

2. **whether_only_child**: 0=no, 1=yes

3. **birth_place**: 0=countryside, 1=town, 2=small_city, 3=medium_to_large_cities

4. **family_economic_status**: 0=extremely_poor, 1=poor, 2=average, 3=good, 4=rich

5. **major**: 0=liberal, 1=science, 2=art

6. **grade**: 0=postgraduate, 1=undergraduate_grade_five, 2=junior, 3=sophomore, 4=freshman, 5=senior

7. **psychiatric_symptoms**: range from 4 to 16

8. **suicide**: range from 4 to 16

9. **dependence**: range from 4 to 16

10. **impulsivity**: range from 4 to 16                                                         

11. **compulsion**: range from 4 to 16

12. **sleeping_disturbance**: range from 4 to 16

13. **internet_addiction**: range from 5 to 20

14. **hostile_aggression**: range from 4 to 16

15. **self_injury_behaviors:** range from 4 to 16

16. **eating_problems**: range from 4 to 16


```{r}
df <- read.csv("data/mental_health_data.csv", )
df <- df[, -1]

# Rename existing columns
colnames(df)[colnames(df) == "whether.only.child"] <- "whether_only_child"
colnames(df)[colnames(df) == "birth.place"] <- "birth_place"
colnames(df)[colnames(df) == "family.economic.status"] <- "family_economic_status"
colnames(df)[colnames(df) == "psychiatric.symptoms"] <- "psychiatric_symptoms"
colnames(df)[colnames(df) == "sleeping.disturbance"] <- "sleeping_disturbance"
colnames(df)[colnames(df) == "internet.addiction"] <- "internet_addiction"
colnames(df)[colnames(df) == "hostile.aggression"] <- "hostile_aggression"
colnames(df)[colnames(df) == "self.injury.behaviors"] <- "self_injury_behaviors"
colnames(df)[colnames(df) == "eating.problems"] <- "eating_problems"

# Transforming variables into factors
df$gender <- factor(df$gender, levels = c(0, 1), labels = c("female", "male"))
df$whether_only_child <- factor(df$whether_only_child, levels = c(0, 1), labels = c("no", "yes"))
df$birth_place <- factor(df$birth_place, levels = 0:3, labels = c("countryside", "town", "small_city", "medium_to_large_cities"))
df$family_economic_status <- factor(df$family_economic_status, levels = 0:4, labels = c("extremely_poor", "poor", "average", "good", "rich"))
df$major <- factor(df$major, levels = 0:2, labels = c("liberal", "science", "art"))
df$grade <- factor(df$grade, levels = 0:5, labels = c("postgraduate", "undergraduate_grade_five", "junior", "sophomore", "freshman", "senior"))
df$psychiatric_symptoms <- factor(df$psychiatric_symptoms)
df$suicide <- factor(df$suicide)
df$dependence <- factor(df$dependence)
df$impulsivity <- factor(df$impulsivity)
df$compulsion <- factor(df$compulsion)
df$sleeping_disturbance <- factor(df$sleeping_disturbance)
df$internet_addiction <- factor(df$internet_addiction)
df$hostile_aggression <- factor(df$hostile_aggression)
df$self_injury_behaviors <- factor(df$self_injury_behaviors)
df$eating_problems <- factor(df$eating_problems)
```

